Make any stupid plot you want

### 2021-02-26 draft/sections/implementation.tex
latex errors + cite tweaks

\begin{document}
\section{Prototype Implementation: Matplottoy}
\label{sec:implementation}
\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/line_1.png}
    \end{subfigure}
    \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
    \label{fig:code_scatter_line}
\end{figure}
To prototype our model, we implemented the artist classes for the scatter and line plots shown in figure~\ref{fig:code_scatter_line} because they differ in every attribute: different visual channels \vchannel\ that composite to different marks \vmark\ with different continuities \vindex\.  We make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations.

To generate the images in figure~\ref{fig:code_scatter_line}, we instantiate \mintinline{python}{fig, ax} artists that will contain the new \mintinline{python}{Point, Line} primitive objects we implemented based on our topology model.

@@

\begin{minted}[escapeinsize=||]{python}
class ArtistClass(matplotlib.artist.Artist):
    |\label{code:artist}|
    def __init__(self, data, transforms, *args, **kwargs):
        # properties that are specific to the graphic but not the channels
        self.data = data 
        self.transforms = transforms
        super().__init__(*args, **kwargs)

@@

The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal. The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that we can use to query back into data to select a subset as discussed in section~\ref{sec:data_sheaf_stalk}. The \vchannel\ functions are then applied to the data to generate the visual section \vsection\ that here is the object \mintinline{python}{visual}. The conversion from data to visual space is simplified here to directly show that it is the encoding \vchannel\ applied to the component. In the full implementation, we allow for fixed visual parameter, such as setting a constant color for all sections, by verifying that the named component is in \dfiber\ before accessing the data. If the data component name is not in \dfiber\, this is interpreted to mean this component is a thickening of \vtotal that could be pulled back to \dtotal\ via an inverse identity \vchannel.

The components of the visual object, denoted by the Python unpacking convention \mintinline{python}{**visual} are then passed into the \mintinline{python}{assembly} function that is \vmarkd. This assembly function is responsible for generating a representation of the glyph such that it could be serialized to recreate a static version of the graphic. Although \mintinline{python}{assemble} could be implemented outside the class such that it returns an object the artist could then parse to set attributes, the attributes are directly set here to reduce indirection. This artist is not optimized because we prioritized demonstrating the separability of \vchannel\ and \vmarkd. The last step in the artist function is handing itself off to the renderer. The extra \mintinline{python}{*arg, **kwargs} arguments in \mintinline{python}{__init__,draw} are artifacts of how these objects are currently implemnted in Matplotlib.

The \mintinline{python}{Point} artist builds on \mintinline{python}{collection} artists because collections are optimized to efficiently draw a sequence of primitive point and area marks. In this prototype, the scatter marker shape is fixed as a circle, and the only visual fiber components are x and y position, size, and the facecolor of the marker. We only show the \mintinline{python}{assembly} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist shown in \ref{code:artist}. 
\begin{minted}{python}
class Point(mcollections.Collection):
    def assemble(self, x, y, s, facecolors='C0' ):
        # construct geometries of the circle glyphs in visual coordinates
        self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
                    for (xi, yi, si) in zip(x, y, s)] 
        # set attributes of glyphs, these are vectorized 
        # circles and facecolors are lists of the same size
        self.set_facecolors(facecolors)
\end{minted} 
The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices, as described in section~\ref{sec:triangulization}. Even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint has corresponding values in section $\dsection(\dbasepoint_{i})$. This means that all the data on one vertex maps to one glyph. To ensure the integrity of the section, \mintinline{python}{view} must be atomic. This means that the values cannot change after the method is called in draw until a new call in draw. We put this constraint on the return of the \mintinline{python}{view} method so that we do not risk race conditions.

@@

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_v.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_h.png}
    \end{subfigure}
    \caption{Frequency of Penguin types visualized as discrete bars. }
    \label{fig:code_bar_simple}
\end{figure}
The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist. The have the same required \vfiber\ components of (position, length), but in \mintinline{pyton}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which controls whether the bars are arranged vertically or horizontally. This parameter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd\.

@@

\end{minted}
The \mintinline{python}{draw} method here has a more complex unpacking of visual encodings to support passing in visual component data directly. This is vastly simplifies building composite objects as the alternative would be higher order functions that take as input the transforms passed in by the user. This construction supports a constant visual parameter, an identity transform where the value is the same in \dtotal\ and \vtotal\, and setting the visual component directly. The \mintinline{python}{assemble} function constructs bars and sets their face and edge colors. The \mintinline{python}{make_bars} function converts the input position and length to the coordinates of a rectangle of the given width. Defaults are provided for 'width' and 'floor' to make this function more reusable. Typically the defaults are used for the type of chart shown in figure~\ref{fig:code_bar_simple}, but these visual variables are often set when building composite versions of this chart type as discussed in section~\ref{sec:code_case_study}.

\subsection{Encoders \vchannel}
\label{sec:code_channels}
As mentioned above, the encoding dictionary is specified by the visual fiber component, the corresponding data fiber component, and the mapping function. The visual parameter serves as the dictionary key because the visual representation is constructed from the encoding applied to the data  $\vsection = \vchannel \circ \dsection$. For the scatter plot, the mappings for the visual fiber components $\vfiber=(x,y, facecolors, s)$ are defined as
\begin{minted}{python}
cmap =  color.Categorical({'true':'deeppink', 'false':'deepskyblue'})
transforms = {'x': {'name': 'v4', 'encoder': lambda x: x},
              'y': {'name': 'v2', 'encoder': lambda x: x},
              'facecolors': {'name':'v3', 'encoder': cmap}, 
              's':{'name': None , 'encoder': lambda _: itertools.repeat(.02)}}
\end{minted}
where the position \textit{(x,y)} \vchannel\ transformers are identity functions. The size \textit{s} transformer is not acting on a component of \dfiber, instead it is a \vchannel\ that returns a constant value. While size could be embedded inside the \mintinline{python}{assembly} function, it is added to the transformers to illustrate user configured visual parameters that could either be constant or mapped to a component in \dfiber. The identity and constant \vchannel\ are explicitly implemented here to demonstrate their implicit role in the visual pipeline, but they are somewhat optimized away in \mintinline{python}{Bar}. More complex encoders can be implemented as callable classes, such as 
\begin{minted}{python}
class Categorical:
    def __init__(self, mapping):
        # check that the conversion is to valid colors
        assert(mcolors.is_color_like(color) for color in mapping.values())
        self._mapping = mapping

@@

where \mintinline{python}{__init__} can validate that the output of the \vchannel\ is a valid element of the \vfiber\ component the \vchannel\ function is targeting. Creating a callable class also provides a simple way to swap out the specific (data, value) mapping without having to reimplement the validation or conversion logic. A test for equivariance can be implemented trivially
\begin{minted}{python}
def test_nominal(values, encoder):
    m1 = list(zip(values, encoder(values)))
    random.shuffle(values)
    m2 = list(zip(values, encoder(values)))
    assert sorted(m1) == sorted(m2)
\end{minted}
but is currently factored out of the artist for clarity. In this example, \mintinline{python}{is_nominal} checks for equivariance of permutation group actions by applying the encoder to a set of values, shuffling values, and checking that (value, encoding) pairs remain the same.

\subsection{Data \dtotal}
\label{sec:code_data}
The data input into the \mintinline{pythin}{Artist} will often be a wrapper class around an existing data structure. This wrapper object must specify the fiber components \dfiber\ and connectivity \dbase\ and have a \mintinline{view} method that returns an atomic object that encapsulates \dsection\. The object returned by the view must be key valued pairs of \mintinline{python}|{component name : component section}| where each section is a component as defined in equation~\ref{eq:data_section_decompose}. To support specifying the fiber bundle, we define a \mintinline{python}{FiberBundle} data class\cite{DataclassesDataClasses}

@@

that asks the user to specify how \dbase is triangulated and the attributes of \dfiber. Python dataclasses are a good abstraction for the fiber bundle class because the \mintinline{python}{FiberBundle} class only stores data. The \dbase\ is specified as tables because the \mintinline{python}{assembly} functions expect tables that match the continuity of the graphic; scatter expects a vertex table because it is discontinuous, line expects an edge table because it is 1D continuous. The fiber informs appropriate choice of \vchannel\, therefore it is a dictionary of attributes of the fiber components.

To generate the scatter plot in figure~\ref{fig:code_scatter_line}, we fully specify a dataset with random keys and values in a section chosen at random form the corresponding fiber component. The fiberbundle \mintinline{python}{FB} is a class level attribute since all instances of \mintinline{python}{code}{VertexSimplex} come from the same fiberbundle.
\begin{minted}{python}
class VertexSimplex: #maybe change name to something else
    """Fiberbundle is consistent across all sections
    """
    FB = FiberBundle({'tables': ['vertex']},  
            {'v1': {'type': float,'monoid':'interval','range': [0,1]},
             'v2': {'type': str, 'monoid':'nominal', 'range':['true', 'false']},
             'v3': {'type': float, 'monoid':'interval', 'range':[2,3]}})

@@

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/linec_1.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/lined_1.png}
    \end{subfigure}
    \caption{Continuous and discontinuous lines as defined by different data models, but generated with the same \vartisteq=\mintinline{python}{artist}}
    \label{fig:circle_step}
\end{figure}

@@

\subsection{Case Study: Penguins}
For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020} since it is multivariate and has a varying number of penguins. We use a version of the data packaged as a pandas dataframe\cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure. The wrapper is very thin because there is explicitly only one section.
\begin{minted}{python}
class DataFrame:
    def __init__(self, dataframe):
        self.FB = FiberBundle(K = {'tables':['vertex']},
                              F = dict(dataframe.dtypes))
        self._tau = dataframe.iloc
        self._view = dataframe

@@

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_stacked.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_grouped.png}
    \end{subfigure}
    \caption{Penguin count disaggregated by island and species}
    \label{fig:code_bar_multi}
\end{figure}

The stacked and grouped bar charts in figure~\ref{fig:code_bar_multi} are both out of \mintinline{python}{Bar} artists such that the difference between \mintinline{Python}{StackedBar} and \mintinline{Python}{GroupedBar} is specific to the ways in which the \mintinline{python}{Bar} are arranged differently. These two artists have identical \mintinline{{python}{__init__} and \mintinline{python}{draw} methods. As with \mintinline{python}{Bar}, the orientation is set on instantiation. In both these artists, we separate the transforms applied to only one component \mintinline{python}{transforms} and the case \mintinline{python}{mtransforms} where the same transform is applied to multiple components such that \votal\ has multiple components that map to the same retinal variable. 
\begin{minted}{python}
class StackedBar(martist.Artist):
    def __init__(self, data, transforms, mtransforms, orientation='v', *args, **kwargs):
        """
        Parameters
        ----------

@@

\end{minted}
Since all the visual transformation is passed through to \mintinline{python}{Bar}, the \mintinline{python}{draw} method does not do any visual transformations. In \mintinline{Python}{StackedBar} the \mintinline{python}{view} is used to adjust the \mintinline{python}{floor} for every subsequent bar chart since a stacked bar chart is bar chart area marks concatenated together in the length parameter. In contrast, \mintinline{Python}{GroupedBar} does not even need the view, but instead keeps track of the relative position of each group of bars in the visual only variable \mintinline{python}{offset}.

@@

        for gid, group in enumerate(self.mtransforms):
            group.update(self.transforms)
            width = group.get('width', .8)
            group['width'] = width/ngroups
            group['offset'] = gid/ngroups*width 
            bar = Bar(self.data, group, self.orientation, transform=self.axes.transData)     
            self.children.append(bar)
\end{minted}{python}

Since the only difference between these two glyphs is in the composition of \mintinline{python}{Bar}, they take in the exact same transform specification dictionaries. \mintinline{python}{transform} dictates the position of the group, in this case by island the penguins are found on.

@@

but differ in assembly \vmarkd. By decomposing the architecture into data, visual encoding, and assembly steps, we are able to build components that are more flexible and also more self contained than the existing code base. While very rough, this API demonstrates that the ideas presented in the math framework are implementable. In choosing a functional approach, if not implementation, we provide a framework for library developers to build reusable encoder \vchannel\, assembly \vmarkd and artists \vartist. We argue that  if these functions are built such that they are equivariant with respect to monoid actions and the graphic topology is a deformation retraction of the data topology, then the artist by definition will be a structure and property preserving map from data to graphic. 
\end{document}

>>>>

\begin{document}
\section{Prototype Implementation: Matplottoy}
\label{sec:implementation}
\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/line_1.png}
    \end{subfigure}
    \caption{Scatter plot and line plot implemented using prototype artists and data models, building on Matplotlib rendering.}
    \label{fig:code_scatter_line}
\end{figure}
To prototype our model, we implemented the artist classes for the scatter and line plots shown in figure~\ref{fig:code_scatter_line} because they differ in every attribute: different visual channels \vchannel\ that composite to different marks \vmark\ with different continuities \vindex\.  We make use of the Matplotlib figure and axes artists \cite{hunterArchitectureOpenSource,hunterMatplotlib2DGraphics2007} so that we can initially focus on the data to graphic transformations. 
To generate the images in figure~\ref{fig:code_scatter_line}, we instantiate \mintinline{python}{fig, ax} artists that will contain the new \mintinline{python}{Point, Line} primitive objects we implemented based on our topology model.

@@

\begin{minted}{python}
class ArtistClass(matplotlib.artist.Artist):
    def __init__(self, data, transforms, *args, **kwargs):
        # properties that are specific to the graphic but not the channels
        self.data = data 
        self.transforms = transforms
        super().__init__(*args, **kwargs)

@@

The data is fetched in section \dsection\ via a \mintinline{python}{view} method on the data because the input to the artist is a section on \dtotal. The \mintinline{python}{view} method takes the \mintinline{python}{axes} attribute because it provides the region in graphic coordinates \gbase\ that we can use to query back into data to select a subset as discussed in section~\ref{sec:artist_sheaf_stalk}. The \vchannel\ functions are then applied to the data to generate the visual section \vsection\ that here is the object \mintinline{python}{visual}. The conversion from data to visual space is simplified here to directly show that it is the encoding \vchannel\ applied to the component. In the full implementation, we allow for fixed visual parameter, such as setting a constant color for all sections, by verifying that the named component is in \dfiber\ before accessing the data. If the data component name is not in \dfiber\, this is interpreted to mean this component is a thickening of \vtotal that could be pulled back to \dtotal\ via an inverse identity \vchannel.

The components of the visual object, denoted by the Python unpacking convention \mintinline{python}{**visual} are then passed into the \mintinline{python}{assemble} function that is \vmarkd. This assembly function is responsible for generating a representation of the glyph such that it could be serialized to recreate a static version of the graphic. Although \mintinline{python}{assemble} could be implemented outside the class such that it returns an object the artist could then parse to set attributes, the attributes are directly set here to reduce indirection. This artist is not optimized because we prioritized demonstrating the separability of \vchannel\ and \vmarkd. The last step in the artist function is handing itself off to the renderer. The extra \mintinline{python}{*arg, **kwargs} arguments in \mintinline{python}{__init__,draw} are artifacts of how these objects are currently implemnted in Matplotlib.

The \mintinline{python}{Point} artist builds on \mintinline{python}{collection} artists because collections are optimized to efficiently draw a sequence of primitive point and area marks. In this prototype, the scatter marker shape is fixed as a circle, and the only visual fiber components are x and y position, size, and the facecolor of the marker. We only show the \mintinline{python}{assemble} function here because the \mintinline{python}{__init__, draw} are identical the prototype artist.
\begin{minted}{python}
class Point(mcollections.Collection):
    def assemble(self, x, y, s, facecolors='C0' ):
        # construct geometries of the circle glyphs in visual coordinates
        self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
                    for (xi, yi, si) in zip(x, y, s)] 
        # set attributes of glyphs, these are vectorized 
        # circles and facecolors are lists of the same size
        self.set_facecolors(facecolors)
\end{minted} 
The \mintinline{python}{view} method repackages the data as a fiber component indexed table of vertices. Even though the \mintinline{python}{view} is fiber indexed, each vertex at an index \dbasepoint has corresponding values in section $\dsection(\dbasepoint_{i})$. This means that all the data on one vertex maps to one glyph. To ensure the integrity of the section, \mintinline{python}{view} must be atomic. This means that the values cannot change after the method is called in draw until a new call in draw. We put this constraint on the return of the \mintinline{python}{view} method so that we do not risk race conditions.

@@

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_v.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_h.png}
    \end{subfigure}
    \caption{Frequency of Penguin types visualized as discrete bars. }
    \label{fig:code_bar_simple}
\end{figure}
The bar charts in figure~\ref{fig:code_bar_simple} are generated with a \mintinline{python}{Bar} artist. The have the same required \vfiber\ components of (position, length), but in \mintinline{python}{__init__} of \mintinline{python}{Bar} an additional parameter is set, \texttt{orientation} which controls whether the bars are arranged vertically or horizontally. This parameter only applies holistically to the graphic and never to individual data parameters, and highlights how the model encourages explicit differentiation between parameters in \vtotal\ and graphic parameters applied directly to \vmarkd.

@@

\end{minted}
The \mintinline{python}{draw} method here has a more complex unpacking of visual encodings to support passing in visual component data directly. This is vastly simplifies building composite objects as the alternative would be higher order functions that take as input the transforms passed in by the user. This construction supports a constant visual parameter, an identity transform where the value is the same in \dtotal\ and \vtotal, and setting the visual component directly. The \mintinline{python}{assemble} function constructs bars and sets their face and edge colors. The \mintinline{python}{make_bars} function converts the input position and length to the coordinates of a rectangle of the given width. Defaults are provided for 'width' and 'floor' to make this function more reusable. Typically the defaults are used for the type of chart shown in figure~\ref{fig:code_bar_simple}, but these visual variables are often set when building composite versions of this chart type as discussed in section~\ref{sec:code_case_study}.

\subsection{Encoders \vchannel}
\label{sec:code_channels}
As mentioned above, the encoding dictionary is specified by the visual fiber component, the corresponding data fiber component, and the mapping function. The visual parameter serves as the dictionary key because the visual representation is constructed from the encoding applied to the data  $\vsection = \vchannel \circ \dsection$. For the scatter plot, the mappings for the visual fiber components $\vfiber=(x,y, facecolors, s)$ are defined as
\begin{minted}{python}
cmap =  color.Categorical({'true':'deeppink', 'false':'deepskyblue'})
transforms = {'x': {'name': 'v4', 'encoder': lambda x: x},
              'y': {'name': 'v2', 'encoder': lambda x: x},
              'facecolors': {'name':'v3', 'encoder': cmap}, 
              's':{'name': None , 'encoder': lambda _: itertools.repeat(.02)}}
\end{minted}
where the position \textit{(x,y)} \vchannel\ transformers are identity functions. The size \textit{s} transformer is not acting on a component of \dfiber, instead it is a \vchannel\ that returns a constant value. While size could be embedded inside the \mintinline{python}{assemble} function, it is added to the transformers to illustrate user configured visual parameters that could either be constant or mapped to a component in \dfiber. The identity and constant \vchannel\ are explicitly implemented here to demonstrate their implicit role in the visual pipeline, but they are somewhat optimized away in \mintinline{python}{Bar}. More complex encoders can be implemented as callable classes, such as 
\begin{minted}{python}
class Categorical:
    def __init__(self, mapping):
        # check that the conversion is to valid colors
        assert(mcolors.is_color_like(color) for color in mapping.values())
        self._mapping = mapping

@@

where \mintinline{python}{__init__} can validate that the output of the \vchannel\ is a valid element of the \vfiber\ component the \vchannel\ function is targeting. Creating a callable class also provides a simple way to swap out the specific (data, value) mapping without having to reimplement the validation or conversion logic. A test for equivariance can be implemented trivially
\begin{minted}{python}
def test_nominal(values, encoder):
    m1 = list(zip(values, encoder(values)))
    random.shuffle(values)
    m2 = list(zip(values, encoder(values)))
    assert sorted(m1) == sorted(m2)
\end{minted}
but is currently factored out of the artist for clarity. In this example, \mintinline{python}{is_nominal} checks for equivariance of permutation group actions by applying the encoder to a set of values, shuffling values, and checking that (value, encoding) pairs remain the same.
\subsection{Data \dtotal}
\label{sec:code_data}
The data input into the \mintinline{python}{Artist} will often be a wrapper class around an existing data structure. This wrapper object must specify the fiber components \dfiber\ and connectivity \dbase\ and have a \mintinline{python}{view} method that returns an atomic object that encapsulates \dsection. The object returned by the view must be key valued pairs of \mintinline{python}|{component name : component section}| where each section is a component as defined in equation~\ref{eq:data_section_decompose}. To support specifying the fiber bundle, we define a \mintinline{python}{FiberBundle} data class\cite{DataclassesDataClasses}

@@

that asks the user to specify how \dbase is triangulated and the attributes of \dfiber. Python dataclasses are a good abstraction for the fiber bundle class because the \mintinline{python}{FiberBundle} class only stores data. The \dbase\ is specified as tables because the \mintinline{python}{assemble} functions expect tables that match the continuity of the graphic; scatter expects a vertex table because it is discontinuous, line expects an edge table because it is 1D continuous. The fiber informs appropriate choice of \vchannel\, therefore it is a dictionary of attributes of the fiber components.

To generate the scatter plot in figure~\ref{fig:code_scatter_line}, we fully specify a dataset with random keys and values in a section chosen at random form the corresponding fiber component. The fiberbundle \mintinline{python}{FB} is a class level attribute since all instances of \mintinline{python}{VertexSimplex} come from the same fiberbundle.
\begin{minted}{python}
class VertexSimplex: #maybe change name to something else
    """Fiberbundle is consistent across all sections
    """
    FB = FiberBundle({'tables': ['vertex']},  
            {'v1': {'type': float,'monoid':'interval','range': [0,1]},
             'v2': {'type': str, 'monoid':'nominal', 'range':['true', 'false']},
             'v3': {'type': float, 'monoid':'interval', 'range':[2,3]}})

@@

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/linec_1.png}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/lined_1.png}
    \end{subfigure}
    \caption{Continuous and discontinuous lines as defined by different data models, but generated with the same \vartisteq \mintinline{python}{Line}}
    \label{fig:circle_step}
\end{figure}

@@

\subsection{Case Study: Penguins}
\label{sec:code_case_study}
For this case study, we use the Palmer Penguins dataset\cite{gormanEcologicalSexualDimorphism2014, horstPalmerpenguinsPalmerArchipelago2020} since it is multivariate and has a varying number of penguins. We use a version of the data packaged as a pandas dataframe\cite{nakhaeeMcnakhaeePalmerpenguins2021} since that is a very commonly used Python labeled data structure. The wrapper is very thin because there is explicitly only one section.
\begin{minted}{python}
class DataFrame:
    def __init__(self, dataframe):
        self.FB = FiberBundle(K = {'tables':['vertex']},
                              F = dict(dataframe.dtypes))
        self._tau = dataframe.iloc
        self._view = dataframe

@@

\begin{figure}[H]
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_stacked.png}
        \caption{}
    \end{subfigure}
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\textwidth]{figures/code/bar_grouped.png}
        \caption{}
    \end{subfigure}
    \caption{Penguin count disaggregated by island and species}
    \label{fig:code_bar_multi}
\end{figure}
The stacked and grouped bar charts in figure~\ref{fig:code_bar_multi} are both out of \mintinline{python}{Bar} artists such that the difference between \mintinline{python}{StackedBar} and \mintinline{python}{GroupedBar} is specific to the ways in which the \mintinline{python}{Bar} are stitched together. These two artists have identical constructors and \mintinline{python}{draw} methods. As with \mintinline{python}{Bar}, the orientation is set in the constructor. In both these artists, we separate the transforms applied to only one component and the case \mintinline{python}{mtransforms} where the same transform is applied to multiple components such that \vtotal\ has multiple components that map to the same retinal variable.

\begin{minted}{python}
class StackedBar(martist.Artist):
    def __init__(self, data, transforms, mtransforms, orientation='v', *args, **kwargs):
        """
        Parameters
        ----------

@@

\end{minted}

Since all the visual transformation is passed through to \mintinline{python}{Bar}, the \mintinline{python}{draw} method does not do any visual transformations. In \mintinline{python}{StackedBar} the \mintinline{python}{view} is used to adjust the \mintinline{python}{floor} for every subsequent bar chart since a stacked bar chart is bar chart area marks concatenated together in the length parameter. In contrast, \mintinline{python}{GroupedBar} does not even need the view, but instead keeps track of the relative position of each group of bars in the visual only variable \mintinline{python}{offset}.

@@

        for gid, group in enumerate(self.mtransforms):
            group.update(self.transforms)
            width = group.get('width', .8)
            group['width'] = width/ngroups
            group['offset'] = gid/ngroups*width 
            bar = Bar(self.data, group, self.orientation, transform=self.axes.transData)     
            self.children.append(bar)
\end{minted}

Since the only difference between these two glyphs is in the composition of \mintinline{python}{Bar}, they take in the exact same transform specification dictionaries. The \mintinline{python}{transform} dictionary dictates the position of the group, in this case by island the penguins are found on.

@@

    but differ in assembly \vmarkd. By decomposing the architecture into data, visual encoding, and assembly steps, we are able to build components that are more flexible and also more self contained than the existing code base. While very rough, this API demonstrates that the ideas presented in the math framework are implementable. In choosing a functional approach, if not implementation, we provide a framework for library developers to build reusable encoder \vchannel\, assembly \vmarkd and artists \vartist. We argue that  if these functions are built such that they are equivariant with respect to monoid actions and the graphic topology is a deformation retraction of the data topology, then the artist by definition will be a structure and property preserving map from data to graphic.

\end{document}

### 2021-02-26 draft/sections/intro.tex
latex errors + cite tweaks

\section{Background}
One of the reasons we developed a new formalism rather than adopting the architecture of an existing library is that most information visualization software design patterns, as categorized by Heer and Agrawala\cite{HeerSoftware2006}, are tuned to very specific data structures. This in turn restricts the design space of visual algorithms that display information (the visualization types the library supports) since the algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighLevel2004}. In proposing a new architecture, we contrast the trade offs libraries make, describe different types of data continuity, and discuss metrics by which a visualization library is traditionally evaluated.

@@

There are also visualization tools where there is no single core structure, and instead internally carry around many different representations of data. Matplotlib, has this structure, as does VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{ramachandranMayaVi2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}. Where GoG and ImageJ type libraries have very consistent APIs for their visualization tools because the data structure is the same, the APIs for visualizations in VTK and Matplotlib are significantly dependent on the structure of the data it expects. This in turn means that every new type of visualization must carry implicit assumptions about data structure in how it interfaces with the input data. This has lead to poor API consistency and brittle code as every visualization type has a very different point of view on how the data is structured. This API choice particularly breaks down when the same dataset is fed into visualizations with different assumptions about structure or into a dashboard consisting of different types of visualization\cite{a.sarikayaWhatWeTalk2019,fewDashboardConfusionRevisited2007} because there is no consistent way to update the data and therefore no consistent way of guaranteeing that the views stay in sync. Our model is a structure dependent formalism, but then also provides a core representation of that structure that is abstract enough to provide a common interface for many different types of visualization.

>>>>

\section{Background}
One of the reasons we developed a new formalism rather than adopting the architecture of an existing library is that most information visualization software design patterns, as categorized by Heer and Agrawala\cite{HeerSoftware2006}, are tuned to very specific data structures. This in turn restricts the design space of visual algorithms that display information (the visualization types the library supports) since the algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{ToryRethinkingVisualization2004}. In proposing a new architecture, we contrast the trade offs libraries make, describe different types of data continuity, and discuss metrics by which a visualization library is traditionally evaluated.

@@

There are also visualization tools where there is no single core structure, and instead internally carry around many different representations of data. Matplotlib, has this structure, as does VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{RamachandranMayaVI2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}. Where GoG and ImageJ type libraries have very consistent APIs for their visualization tools because the data structure is the same, the APIs for visualizations in VTK and Matplotlib are significantly dependent on the structure of the data it expects. This in turn means that every new type of visualization must carry implicit assumptions about data structure in how it interfaces with the input data. This has lead to poor API consistency and brittle code as every visualization type has a very different point of view on how the data is structured. This API choice particularly breaks down when the same dataset is fed into visualizations with different assumptions about structure or into a dashboard consisting of different types of visualization\cite{a.sarikayaWhatWeTalk2019,fewDashboardConfusionRevisited2007} because there is no consistent way to update the data and therefore no consistent way of guaranteeing that the views stay in sync. Our model is a structure dependent formalism, but then also provides a core representation of that structure that is abstract enough to provide a common interface for many different types of visualization.
